\newcommand{\MAIN}{MultiSortedAlgebra}
%
\newcommand{\TITLE}{%
  Birkhoff's Completeness Theorem for Multi-Sorted Algebras
  Formalized in Agda
}
\newcommand{\DATE}{3 June 2021}
%
\newcommand{\ABSTRACT}{
  This document provides a formal proof of Birkhoff's completeness
  theorem for multi-sorted algebras which states that any equational
  entailment valid in all models is also provable in the equational
  theory.
  More precisely, if a certain equation is valid in all models that
  validate a fixed set of equations, then this equation is derivable
  from that set using the proof rules for a congruence.

  The proof has been formalized in Agda version 2.6.2 with the
  Agda Standard Library version 1.7 and this
  document reproduces the commented Agda code.
}
%
\newcommand{\INTRO}{

  \section{Introduction}

  Birkhoff's completeness theorem \citeyearpar{birkhoff:1935}
  has been formalized in type theory before, even in Agda
  \citep[Thm~.3.1]{guntherGadeaPagano:lsfa17}.
  Our formalization makes the following decisions:

  \begin{enumerate}

  \item We use indexed containers
    \citep{altenkirchGhaniHancockMcBrideMorris:jfp15}
    aka Peterson-Synek (interaction) trees.
    Given a set $S$ of sort symbols, a signature over $S$ is an
    indexed endo-container, which has three components:
    \begin{enumerate}

    \item Per sort $s : S$, a set $O_s$ of operator symbols.
      (In the container terminology, these are called
      \emph{shapes} for index $s$, and in the interaction tree terminology,
      \emph{commands} for state $s$.)

    \item Per operator $o : O_s$, a set $A_o$, the arity of operator
      $o$.  The arity is the index set for the arguments of the
      operator, which are then given by a function over $A_o$.
      (In the other terminologies, these are the \emph{positions} or
      \emph{responses}, resp.)

    \item Per argument index $i : A_o$, a sort $s_i : S$ which
      denotes the sort of the $i$th argument of operator $o$.
      (In the interaction tree terminology, this is the \emph{next} state.)

    \end{enumerate}

    Closed terms of a multi-sorted algebra
    (aka first-order terms) are then
    concrete interaction trees, i.e., elements of the indexed $W$-type
    pertaining to the container.

    Note that all the ``set''s we mentioned above come with a size,
    see next point.

  \item Universe-polymorphic:
    As we are working in a predicative and constructive
    meta-theory, we have to be aware of the size
    (i.e., inaccessible cardinality) of the sets.
    Our formalization is universe-polymorphic to ensure good
    generality, resting on the universe-polymorphic
    Agda Standard Library.

    In particular, there is no such thing as ``all models''; rather we
    can only quantify over models of a certain maximum size.
    The completeness theorem consequently does not require validity of
    an entailment in all models, but only in all models of a certain
    size, which is given by the size of the generic model, i.e., the
    term model.  The size of the term model in turn is determined by
    the size of the signature of the multi-sorted algebra.

  \item Open terms (with free variables) are obtained as the
    \emph{free monad} over the container.  Concretely, we make a new
    container that has additional nullary operator symbols, which
    stand for the variables.
    Terms are intrinsically typed, i.e., the set of terms is actually
    a family of sets indexed by a sort and a context of sorted
    variables in scope.

  \item No lists:
    We have no finiteness restrictions whatsover, neither the number
    of operators need to be finite, nor the number of arguments of an
    operator, nor the set of variables that are in scope of a term.
    (Note however, since terms are finite trees, they can actually
    mention only a finite number of variables from the possibly
    infinite supply.)

  \end{enumerate}
}
%
\newcommand{\CONCLUSION}{%
  Q.E.D.

  \section{Related work}

  \cite{guntherGadeaPagano:lsfa17} further formalize signature
  morphisms.  These would be, in our setting, morphisms of indexed
  containers, described by
  \cite{altenkirchGhaniHancockMcBrideMorris:jfp15}, albeit in a
  slightly different semantics, slice categories.

  DeMeo's rather comprehensive development \citeyearpar{deMeo:ualib-arxiv}
  formalizes single-sorted algebras up to the
  Birkhoff's HSP theorem in Agda.  DeMeo's signatures are containers;
  even though he does not make this connection explicit, it inspired
  the use of indexed containers in the present development.
  DeMeo's formalization is basis for
  \url{https://github.com/ualib/agda-algebras}.

  \cite{amatoMaggesiBrogi:univ-alg-arxiv-2021} formalize multi-sorted
  algebras with finitary operators in UniMath.  Limiting to
  finitary operators is due to the
  restrictions of the UniMath type theory, which does not have W-types
  nor user-defined inductive types.  These restrictions also prompt
  the authors to code terms as lists of stack machine instructions
  rather than trees.

  \cite{lyngeSpitters:types19} formalize multi-sorted algebras in
  HoTT, also restricting to finitary operators.  Using HoTT they can
  define quotients as types, obsoleting setoids.
  They prove three
  isomorphism theorems concerning sub- and quotient algebras.  A
  universal algebra or varieties are not formalized.

  \paragraph*{Acknowledgments.}
  Standing on the shoulders of giants: the present concise formalization is
  enabled by the well-organized standard library of Agda maintained by
  Matthew Daggitt and Guillaume Allais.
  Thanks to Jacques Carette for discussions and constructive review of
  the Agda code.

  \GENERATED
%
\bibliographystyle{abbrvnat}
\bibliography{auto-MultiSortedAlgebra}
}

\input{template}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
